\documentclass[12 pt, leqno]{article}
\usepackage{latexsym}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{fancyhdr}
\usepackage{graphicx}
\usepackage[margin=1in]{geometry}
\setlength\parindent{20pt}
\usepackage{setspace} 
\newcommand{\indep}{\rotatebox[origin=c]{90}{$\models$}}


\begin{document}

\title{Extra things from SOC 351}
\author{Siddhartha Basu}
\date{\today}
\maketitle


\section{Things left over}

Diff in diff assumptions

Regression discontinuity assumptions

Advanced IV (Mostly harmless chapter on IV)

Advanced fixed effects (Wooldridge chapter on FE, really getting non-nested multi-dimensional FE)




\section{Questions for Florencia}
\begin{itemize}
\item Does exogeneity restriction only matter for $D$, or other $X$? Answer: D is the most important variable. If you have a violation for other X, it is not world ending...however this makes it more important for you to have linearity. 
\item Strict exogeneity $cov(x_{it}, e_{is}) = 0$ for all $t, s$ vs. no time varying confounders. 
\item Going over fixed effects assumptions (treatment random, strict exo, treatment on outcome)
\item Exogeneity $cov(x,e) = 0$ vs. $E(e_{i1|d=1} - e_{i1|d = 0})$
\item Variation when you have multiple FE (host, month), how do assumptions change?
\item Example: variation of birthweight within twins to predict causal impact of birthweight on test score
\item Example: individual fixed effect for effect on marriage on wages
\item Example: school and year fixed effects, then regress failure rate on murder in municipality
\item Example: listingxchannel and marketxmonth fixed effects for host heterogeneity
\item Example: route fixed effects for causal impact of airline on delay?
\item Example: Moorish influence on Spanish government?
\end{itemize}

New questions for florencia:
\begin{itemize}
\item `within individual variation to identify' -- How does this change for multi-dimensional fixed effects?
\item Do you need to assume that `marriage is an absorbing state' for deviation from means estimator to work?
\item General intuition for non-nested fixed effects models. 
\item How are the $\beta$'s the same for the deviation from means estimator or first difference estimator?
\item Does clustering solve serial correlation problem if you have year fixed effects?
\item ask for slides
\end{itemize}


\end{document}